\thetitle

### 2020-10-28 sun_phases/spin_qudit_tomo.tex
minor wording changes

\begin{abstract}
  We consider the task of performing quantum state tomography on a $d$-state spin qudit, using only measurements of spin projection onto different quantization axes.
  By mapping this quantum tomography problem onto the classical problem of signal recovery on the sphere, we prove that full reconstruction of arbitrary mixed states is possible, and requires a minimum number of measurement axes, $r_d$, that is bounded by $2d-1\le r_d\le d^2$.
  We conjecture that $r_d=2d-1$, which we have verified for all $d\le120$ using numerical experiments.
  We then provide an algorithm with $O\p{r^3}$ runtime for certifying strict upper bounds on the error with which measurements of spin projection along $r$ measurement axes can be used to reconstruct an unknown quantum state.
  Our algorithm motivates a simple randomized tomography protocol, which we use to study the trade-off between the number of measurement axes and state reconstruction error [\red{note: this part is not yet complete}].
  [\red{todo: maybe look into a simplified tomography procedure for pure states?]}
\end{abstract}

@@

It will be useful to think of rotation matrix elements $D^\ell_{mn}$ and $\D^\ell_{mn}$ as scalar functions of Euler angles $\v\omega\in\Omega_3$.
As functions of $\v\omega$, the fundamental rotation matrix elements $D_{mn}^\ell$ satisfy the orthogonality relations \cite{brown2003rotational}
\begin{align}
  \int_{\Omega_3} \d\Omega_3\p{\v\omega}
  D^{\ell}_{mn}\p{\v\omega}^* D^{\ell'}_{m'n'}\p{\v\omega}
  = \f{8\pi^2}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'} \delta_{nn'},
  \label{eq:ortho_full}
\end{align}
where integration is performed over all $\v\omega\in\Omega_3\simeq[0,2\pi]\times[0,\pi]\times[0,2\pi]$ with integration measure $\d\Omega_3\p{\v\omega} \equiv \d\alpha_{\v\omega} \d\beta_{\v\omega} \d\gamma_{\v\omega} \sin\beta_{\v\omega}$.
In particular, the restriction of $D^\ell_{m,0}$ to the sphere $\Omega_2$ at $\gamma_{\v\omega}=0$ recovers (up to complex conjugation) the spherical harmonics $Y_{\ell m}$, with orthogonality relations
\begin{align}
  \int_{\Omega_2} \d\Omega_2\p{\v v}
  D^{\ell}_{m,0}\p{\v v}^* D^{\ell'}_{m',0}\p{\v v}
  = \f{4\pi}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'},
  \label{eq:ortho_sphere}
\end{align}
where integration is performed over all $\v v\in\Omega_2\simeq[0,2\pi]\times[0,\pi]$ with integration measure $\d\Omega_2\p{\v v} \equiv \d\alpha_{\v v} \d\beta_{\v v} \sin\beta_{\v v}$, and for ease of notation we implicitly associate doublets $\v v=\p{\alpha_{\v v},\beta_{\v v}}$ with the triplet $\p{\alpha_{\v v},\beta_{\v v},0}$, so that for example $R\p{\v v} = R\p{\alpha_{\v v},\beta_{\v v},0}$ and $T_{\v v\ell m} = T_{\p{\alpha_{\v v},\beta_{\v v},0},\ell m}$.
The same orthogonality relations in Eqs.~\eqref{eq:ortho_full} and \eqref{eq:ortho_sphere} are inherited by the rotation matrix elements $\D_{mn}^\ell$ through the fact $\R\p{\v\omega}$ is the adjoint action of $R\p{\v\omega}$ (see Appendix \ref{sec:ortho}), and the functions $\set{\D_{m,0}^\ell}$ restricted to $\Omega_2$ are similarly another orthogonal basis for the space $L^2\p{\Omega_2}$ of square-integrable functions on the sphere.
[\red{todo: make figures visualizing $Y_{\ell m}$ and $\D^\ell_{m,0}$ for comparison}]

@@

Moreover, for any fixed degree $\ell$, finding solutions to the reduced recovery problem in Eq.~\eqref{eq:tomo_recovery} is equivalent to the recovery of a function in the degree-$\ell$ subspace of $L^2\p{\Omega_2}$, which is possible with $\abs{V}=2\ell+1$ samples (equal to the dimension of the degree-$\ell$ subspace) \cite{freeden2008spherical}.
In the case of spin qudit tomography, the degree $\ell$ takes a maximal value of $\ell_{\t{max}}\equiv d-1$, so state recovery requires at least as many measurement axes as there are transition operators with degree $\ell_{\t{max}}$, namely $2\ell_{\t{max}}+1=2d-1$.
Since
\begin{enumerate*}
\item there are only $2\ell+1\le2d-1$ transition operators of degree $\ell\le\ell_{\t{max}}$, and
\item measuring spin projections along any axis simultaneously provides measurement data for transition operators of {\it all} degrees $\ell\le\ell_{\t{max}}$,
\end{enumerate*}
one might expect that $2d-1$ suitably chosen measurement axes would always suffice to perform full tomography of a $d$-level spin qudit.
In the language of classical signal recovery, we would expect there to exist a set of $2\ell_{\t{max}}+1$ points on the sphere that enable the recovery of any degree-$\ell$ function with $\ell\le\ell_{\t{max}}$ (given prior knowledge of $\ell$) \footnote{We could further conjecture that if {\it any} set of points $V\subset\Omega_2$ enables the recovery of degree-$\ell_{\t{max}}$ functions on the sphere, then $V$ would allow recovery degree-$\ell$ functions with $\ell\le\ell_{\t{max}}$.  However, this condition is stronger than that necessary to guarantee full recovery of a $d$-level spin qudit state with $2d-1$ measurement axes.}.
Unfortunately, we have no general proof of this conjecture, which may require a deeper analysis of fixed-degree signal recovery on the sphere.
Nonetheless, we have used the tomography protocol in the next section to numerically verify that $2d-1$ measurement axes suffice for all $d\le120$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Optimized tomography protocols and reconstruction error bounds}
\label{sec:protocol}

@@

where we neglect cross terms with different degrees $\ell$ because the corresponding inner products $\obk{T_{\v v\ell,0}|T_{\v v'\ell',0}}=0$ for all $\ell\ne\ell'$.

@@

Indexing these eigenvalues $G^V_k$ and the corresponding (normalized) eigenvectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,\abs{V}d}$, we can construct the orthonormal operators

@@

where for shorthand we use a combined index $j\equiv\p{\v v_j,\ell_j}$ to specify both a measurement axis $\v v_j$ and a degree $\ell_j$, which identify the transition operator $T_j\equiv T_{\v v_j\ell_j,0}$.

@@

The variance in the error with which $\tilde\rho^V_k$ approximates $\rho^V_k$ is then

@@

  &= \Bbk{\p{\tilde\rho^V_k-\rho^V_k}^\dag \p{\tilde\rho^V_k-\rho^V_k}} \\

@@

is a norm uniquely determined by the set of measurement axes $V$.

@@

The bound in Eq.~\eqref{eq:bound} motivates the following simple protocol for spin qudit tomography: select a random set of measurement axes $V$ by uniformly sampling points on the sphere \footnote{To sample a point $\p{\alpha,\beta}$ from the uniform distribution on the sphere (with azimuthal angle $\alpha$ and polar angle $\beta$), you can sample a point $\p{a,b}\in[0,1]\times[0,1]$ from the uniform distribution on the unit square, then set $\alpha=2\pi a$ and $\beta=\arccos\p{1-2b}$.}, and variationally optimize the $2\abs{V}$ parameters in $V$ (two angles for each point $\v v\in V$) to minimize the error norm $\N_V$.
If $\abs{V}$ is too large for variational optimization to be practical, one can instead generate many random sets of measurement axes, and then choose the set that minimizes $\N_V$.
After choosing a set of measurement axes $V$, experimentally characterize the expectation values $\bk{T_j}_\rho$.
If the experimental measurements $E_j$ of $\bk{T_j}_\rho$ are accurate to within an error $\lesssim\epsilon$, then the state $\tilde\rho$ defined in Eq.~\eqref{eq:reconstructed_state} is an approximation of the true state $\rho$ to within a 2-norm distance of $\lesssim\epsilon\N_V$.
Furthermore, if the empirical estimate $\tilde\rho$ has negative eigenvalues, its 2-norm distance from $\rho$ can be further reduced with maximum-likelihood corrections \cite{smolin2012efficient}.
In terms of classical computing costs before and after estimating expectation values $\bk{T_j}_\rho$, this tomography protocol is bottlenecked by the need to diagonalize the overlap matrix $G_V$, which has $d$ blocks of size $\abs{V}\times\abs{V}$.
Altogether, this protocol therefore has an $O(d\abs{V}^3)$ serial or $O(\abs{V}^3)$ parallel runtime.

@@

As a last point, we mention that the cost function $\N_V$ can also evaluate the quality of measurement axes chosen using existing methods for sparse signal recovery on the sphere \cite{rauhut2011sparse, alem2012sparse}.
In practice, we find that uniform random sampling outperforms these methods, in the sense that it generates low-cost sets of measurement axes with higher probability.
Two observations may explain the poor performance of the sparse signal recovery methods.
First, spin qudit tomography essentially boils down to recovering sparse signals with a particular harmonic structure, namely with a fixed degree $\ell$.
The existing sparse signal recovery algorithms do not assume or account for any harmonic structure, and may therefore perform worse than expected on this particular subset of sparse signals.
Second, spin qudit tomography requires recovering signals that are band-limited at degree $d$, but have $O(d)$ non-zero harmonic coefficients (out of $d^2$ coefficients in total).
The existing sparse signal recovery methods, on the other hand, may only perform well for signals with sparsity much smaller than $d$.
In any case, the fact that uniform sampling outperforms these methods only simplifies our randomized tomography protocol.

@@

As functions of Euler angle triplets $\v\omega=\p{\alpha_{\v\omega},\beta_{\v\omega},\gamma_{\v\omega}}$, the fundamental (Wigner) rotation matrix elements $D_{mn}^\ell$ and satisfy the orthogonality relations \cite{brown2003rotational}
\begin{align}
  \int_{\Omega_3} \d\Omega_3\p{\v\omega}
  D^{\ell}_{mn}\p{\v\omega}^* D^{\ell'}_{m'n'}\p{\v\omega}
  = \f{8\pi^2}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'} \delta_{nn'},
  \label{eq:orthogonal}
\end{align}
where integration is performed over all $\v\omega\in\Omega_3\simeq[0,2\pi]\times[0,\pi]\times[0,2\pi]$ with integration measure $\d\Omega_3\p{\v\omega}\equiv\d\alpha_{\v\omega} \d\beta_{\v\omega} \d\gamma_{\v\omega} \sin\beta_{\v\omega}$.
This orthogonality relation is a consequence of the fact that, for a fixed degree $\ell$, the matrix
\begin{align}
  D^\ell \equiv \sum_{m,n} D^\ell_{mn} \op{m}{n}
\end{align}
is an irreducible unitary matrix representation of SO(3), where by unitary we mean that
\begin{align}
  \sum_k D^\ell_{km}\p{\v\omega}^* D^\ell_{kn}\p{\v\omega} = \delta_{mn}
  ~\t{for all}~\v\omega\in\Omega_3\simeq\SO(3).
\end{align}
Unitarity of $D^\ell$ follows from the fact that the matrix $D^\ell\p{\v\omega}$ is essentially the degree-$\ell$ block of the block-diagonal fundamental rotation matrix $R\p{\v\omega}$, which is itself unitary.
The Schur orthogonality relations for irreducible unitary representations of compact Lie groups then imply that \cite{cornwell1997group}
\begin{align}
  \int_{\Omega_3} \d\Omega_3\p{\v\omega}
  D^{\ell}_{mn}\p{\v\omega}^* D^{\ell'}_{m'n'}\p{\v\omega}
  = \f{\abs{\Omega_3}}{\dim\p{D^\ell}}
  \delta_{\ell\ell'} \delta_{mm'} \delta_{nn'},
  \label{eq:ortho_full_apndx}
\end{align}
where $\dim\p{D^\ell}=2\ell+1$ is the dimension of $D^\ell$, and analogously
\begin{align}
  \abs{\Omega_3} \equiv \int_{\Omega_3} d\Omega_3\p{\v\omega} = 8\pi^2.
\end{align}
Orthogonality relations for adjoint matrix elements $\D^\ell_{mn}$ are derived identically from the fact that the matrix
\begin{align}
  \D^\ell \equiv \sum_{m,n} \D^\ell_{mn} \op{m}{n}
\end{align}
is an irreducible unitary representation of SO(3).
This property of $\D^\ell$ follows from the fact that it is an adjoint representation of SO(3), which is isomorphic to the standard representation $D^\ell$ \cite{hall2015lie}.

The decomposition $D^\ell_{mn}\p{\v\omega} = e^{-\i\alpha_{\v\omega}m} d^\ell_{mn}\p{\beta_{\v\omega}} e^{-\i\gamma_{\v\omega}n}$, where $d^\ell_{mn}\p{\beta}\equiv D^\ell_{mn}\p{0,\beta,0}$ is
a matrix element of a reduced fundamental (Wigner) rotation matrix \cite{rose1957elementary}, implies that setting $n'=n$ in in Eq.~\eqref{eq:ortho_full_apndx} makes the integrand independent of $\gamma_{\v\omega}$.
Evaluating the integral over $\gamma_{\v\omega}$ and dividing through by $2\pi$ then recovers the orthogonality relations
\begin{align}
  \int_{\Omega_2} \d\Omega_2\p{\v v}
  D^{\ell}_{mn}\p{\v v}^* D^{\ell'}_{m'n}\p{\v v}
  = \f{4\pi}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'},
\end{align}
where integration is performed over all $\v v\in\Omega_2\simeq[0,2\pi]\times[0,\pi]$ with integration measure $\d\Omega_2\p{\v v} \equiv \d\alpha_{\v v} \d\beta_{\v v} \sin\beta_{\v v}$.
A similar decomposition of the adjoint rotation matrix elements $\D^\ell_{mn}\p{\v\omega}$ (see Appendix \ref{sec:rotations}) implies the identical orthogonality relations.

@@

Here we show how one solution to the problem of signal recovery on the sphere with a particular basis of functions (such as the spherical harmonics $Y_{\ell m}$) can be translated into a solution with a different basis (such as the adjoint rotation matrix elements $\D^\ell_{m,0}$).
Consider a square-integrable function on the sphere, $f\in L^2\p{\Omega_2}$, with the harmonic expansion
\begin{align}
  f\p{\v v} = \sum_{\ell,m} f_{\ell m} B_{\ell m},
\end{align}
where $f_{\ell m}\in\CC$ is a coefficient and $\set{B_{\ell m}}$ is a complete basis for $L^2\p{\Omega_2}$.
Without loss of generality, we will assume that for a fixed degree $\ell$ the functions $\set{B_{\ell m}}$ span the degree-$\ell$ subspace of $L^2\p{\Omega_2}$, and that these functions satisfy orthogonality relations analogous to that in Eq.~\eqref{eq:ortho_sphere}.
A solution to the signal recovery problem amounts to a set of points $V=\set{\v v}\subset\Omega_2$ and coefficients $C_{\ell m}\p{\v v}$ for which
\begin{align}
  f_{\ell m} = \sum_{\v v} C_{\ell m}\p{\v v} f\p{\v v}.
  \label{eq:recovery_solution}
\end{align}
Consider now a different basis $\set{\tilde B_{\ell m}}$ for $L^2\p{\Omega_2}$, with which
\begin{align}
  f\p{\v v} = \sum_{\ell, m} \tilde f_{\ell m} \tilde B_{\ell m},
\end{align}
where we make the same general assumptions about $\set{\tilde B_{\ell m}}$ that we made for $\set{B_{\ell m}}$.
Substituting the harmonic expansion of $f\p{\v v}$ in the original basis $\set{B_{\ell m}}$, we find that
\begin{align}
  \tilde f_{\ell m} = \int_{\Omega_2} \d\Omega_2\p{\v v}
  \tilde B_{\ell m}\p{\v v}^* f\p{\v v}
  = \sum_n G^\ell_{mn} f_{\ell n},
  &&
  G^\ell_{mn} \equiv \int_{\Omega_2} \d\Omega_2\p{\v v}
  \tilde B_{\ell m}\p{\v v}^* B_{\ell n}\p{\v v},
  \label{eq:basis_change}
\end{align}
where integration is performed over all $\v v\in\Omega_2\simeq[0,2\pi]\times[0,\pi]$ with integration measure $\d\Omega_2\p{\v v} \equiv \d\alpha_{\v v} \d\beta_{\v v} \sin\beta_{\v v}$.
Note that the inner product between functions $\tilde B_{\ell m}\p{\v v}$ and $B_{kn}\p{\v v}$ for $k\ne\ell$ vanishes because these functions live in orthogonal subspaces of $L^2\p{\Omega_2}$ \cite{freeden2008spherical}.
Substituting Eq.~\eqref{eq:recovery_solution} into Eq.~\eqref{eq:basis_change}, we find that
\begin{align}
  \tilde f_{\ell m}
  = \sum_n G^\ell_{mn} \sum_{\v v} C_{\ell n}\p{\v v} f\p{\v v}
  = \sum_{\v v} \tilde C_{\ell m}\p{\v v} f\p{\v v},
  &&
  \tilde C_{\ell m}\p{\v v} \equiv \sum_n G^\ell_{mn} C_{\ell n}\p{\v v}.
\end{align}
The coefficients $\tilde C_{\ell m}\p{\v v}$ thus solve the signal recovery problem with the new basis $\set{\tilde B_{\ell m}}$ for $L^2\p{\Omega_2}$.

>>>>

\begin{abstract}
  We consider the task of performing quantum state tomography on a $d$-state spin qudit, using only measurements of spin projection onto different quantization axes.
  By mapping this quantum tomography problem onto the classical problem of signal recovery on the sphere, we prove that full reconstruction of arbitrary mixed states is possible, and requires a minimum number of measurement axes, $r_d$, that is bounded by $2d-1\le r_d\le d^2$.
  We conjecture that $r_d=2d-1$, which we have verified for all $d\le150$ using numerical experiments.
  We then provide a parallelized algorithm with $O\p{r^3}$ runtime for certifying upper bounds on the error with which measurements of spin projection along $r$ quantization axes can be used to reconstruct an unknown quantum state.
  Our algorithm motivates a simple randomized tomography protocol, which we use to study the trade-off between the number of measurement axes and state reconstruction error at a fixed total number of measurements [\red{note: this part is not yet complete}].
  [\red{todo: maybe look into a simplified tomography procedure for pure states?]}
\end{abstract}

@@

It will be useful to think of rotation matrix elements $D^\ell_{mn}$ and $\D^\ell_{mn}$ as scalar functions of Euler angles $\v\omega\in\Omega_3$.
As functions of $\v\omega$, the fundamental rotation matrix elements $D_{mn}^\ell$ satisfy the orthogonality relations \cite{brown2003rotational}
\begin{align}
  \int_{\Omega_3} \d\Omega_3\p{\v\omega}
  D^{\ell}_{mn}\p{\v\omega}^* D^{\ell'}_{m'n'}\p{\v\omega}
  = \f{8\pi^2}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'} \delta_{nn'},
  \label{eq:ortho_full}
\end{align}
where integration is performed over all $\v\omega\in\Omega_3\simeq[0,2\pi]\times[0,\pi]\times[0,2\pi]$ with integration measure $\d\Omega_3\p{\v\omega} \equiv \d\alpha_{\v\omega} \d\beta_{\v\omega} \d\gamma_{\v\omega} \sin\beta_{\v\omega}$.
In particular, the restriction of $D^\ell_{m,0}$ to the sphere $\Omega_2$ at $\gamma_{\v\omega}=0$ recovers (up to complex conjugation) the spherical harmonics $Y_{\ell m}$, with orthogonality relations
\begin{align}
  \int_{\Omega_2} \d\Omega_2\p{\v v}
  D^{\ell}_{m,0}\p{\v v}^* D^{\ell'}_{m',0}\p{\v v}
  = \f{4\pi}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'},
  \label{eq:ortho_sphere}
\end{align}
where integration is performed over all $\v v\in\Omega_2\simeq[0,2\pi]\times[0,\pi]$ with integration measure $\d\Omega_2\p{\v v} \equiv \d\alpha_{\v v} \d\beta_{\v v} \sin\beta_{\v v}$, and for ease of notation we implicitly associate doublets $\v v=\p{\alpha_{\v v},\beta_{\v v}}$ with the triplet $\p{\alpha_{\v v},\beta_{\v v},0}$, so that for example $R\p{\v v} = R\p{\alpha_{\v v},\beta_{\v v},0}$ and $T_{\v v\ell m} = T_{\p{\alpha_{\v v},\beta_{\v v},0},\ell m}$.
The same orthogonality relations in Eqs.~\eqref{eq:ortho_full} and \eqref{eq:ortho_sphere} are inherited by the adjoint rotation matrix elements $\D_{mn}^\ell$ through the fact $\R\p{\v\omega}$ is the adjoint action of $R\p{\v\omega}$ (see Appendix \ref{sec:ortho}), and the functions $\set{\D_{m,0}^\ell}$ restricted to $\Omega_2$ are similarly another orthogonal basis for the space $L^2\p{\Omega_2}$ of square-integrable functions on the sphere.
[\red{todo: make figures visualizing $Y_{\ell m}$ and $\D^\ell_{m,0}$ for comparison}]

@@

Moreover, for any fixed degree $\ell$, finding solutions to the reduced recovery problem in Eq.~\eqref{eq:tomo_recovery} is equivalent to the recovery of a function in the degree-$\ell$ subspace of $L^2\p{\Omega_2}$, which is possible with $\abs{V}=2\ell+1$ samples (equal to the dimension of the degree-$\ell$ subspace) \cite{freeden2008spherical}.
In the case of spin qudit tomography, the degree $\ell$ takes a maximal value of $\ell_{\t{max}}\equiv d-1$, so state recovery requires at least as many measurement axes as there are transition operators with degree $\ell_{\t{max}}$, namely $2\ell_{\t{max}}+1=2d-1$.
Since
\begin{enumerate*}
\item there are only $2\ell+1\le2d-1$ transition operators of degree $\ell\le\ell_{\t{max}}$, and
\item measuring spin projections along any axis simultaneously provides measurement data for transition operators of {\it all} degrees $\ell\le\ell_{\t{max}}$,
\end{enumerate*}
one might expect that $2d-1$ suitably chosen measurement axes would always suffice to perform full tomography of a $d$-level spin qudit.
In the language of classical signal recovery, we would expect there to exist a set of $2\ell_{\t{max}}+1$ points on the sphere that enable the recovery of any degree-$\ell$ function with $\ell\le\ell_{\t{max}}$ (given prior knowledge of $\ell$) \footnote{We could further conjecture that if {\it any} set of points $V\subset\Omega_2$ enables the recovery of degree-$\ell_{\t{max}}$ functions on the sphere, then $V$ would allow recovery degree-$\ell$ functions with $\ell\le\ell_{\t{max}}$.  However, this condition is stronger than that necessary to guarantee full recovery of a $d$-level spin qudit state with $2d-1$ measurement axes.}.
Unfortunately, we have no general proof of this conjecture, which may require a deeper analysis of fixed-degree signal recovery on the sphere.
Nonetheless, we have used the tomography protocol in the next section to numerically verify that $2d-1$ measurement axes suffice for all $d\le150$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reconstruction error bounds}
\label{sec:protocol}

@@

where we neglect cross terms with different degrees $\ell$ because the corresponding inner products vanish, i.e.~$\obk{T_{\v v\ell,0}|T_{\v v'\ell',0}}=0$ for all $\ell\ne\ell'$.

@@

Indexing these eigenvalues $G^V_k$ and the corresponding (normalized) eigenvectors $\v x^V_k\equiv(x^V_{k,1},x^V_{k,2},\cdots)$ by an integer $k\in\set{1,2,\cdots,d^2}$, we can construct the orthonormal operators

@@

where for shorthand we use a combined index $j=\p{\v v_j,\ell_j}$ to specify both a measurement axis $\v v_j$ and a degree $\ell_j$, which identify the transition operator $T_j\equiv T_{\v v_j\ell_j,0}$.

@@

The mean squared error with which $\tilde\rho^V_k$ approximates $\rho^V_k$ is then

@@

  &= \Bbk{\p{\tilde\rho^V_k-\rho^V_k}^*\p{\tilde\rho^V_k-\rho^V_k}} \\

@@

is a norm determined by the set of measurement axes $V$.

@@

The bound in Eq.~\eqref{eq:bound} motivates the following simple protocol for spin qudit tomography: select a random set of measurement axes $V$ by uniformly sampling points on the sphere \footnote{To sample a point $\p{\alpha,\beta}$ from the uniform distribution on the sphere (with azimuthal angle $\alpha$ and polar angle $\beta$), you can sample a point $\p{a,b}\in[0,1]\times[0,1]$ from the uniform distribution on the unit square, and then set $\alpha=2\pi a$ and $\beta=\arccos\p{1-2b}$.}, and variationally optimize the $2\abs{V}$ parameters in $V$ (two angles for each point $\v v\in V$) to minimize the error norm $\N_V$.
If $\abs{V}$ is too large for variational optimization to be practical, one can instead generate many random sets of measurement axes, and then choose the set that minimizes $\N_V$.
After choosing a set of measurement axes $V$, experimentally characterize the expectation values $\bk{T_j}_\rho$.
If the experimental measurements $E_j$ of $\bk{T_j}_\rho$ are accurate to within an error $\lesssim\epsilon$, then the state $\tilde\rho$ defined in Eq.~\eqref{eq:reconstructed_state} is an approximation of the true state $\rho$ to within a 2-norm distance of $\lesssim\epsilon\N_V$.
Furthermore, if the empirical estimate $\tilde\rho$ has negative eigenvalues, its 2-norm distance from $\rho$ can be further reduced with maximum-likelihood corrections \cite{smolin2012efficient}.
In terms of classical computing costs before and after estimating expectation values $\bk{T_j}_\rho$, this tomography protocol is bottlenecked by the need to diagonalize the overlap matrix $G_V$, which has $d$ blocks of size $\abs{V}\times\abs{V}$.
Altogether, this protocol therefore has an $O(d\abs{V}^3)$ serial or $O(\abs{V}^3)$ parallel runtime.

@@

As a last point, we mention that the cost function $\N_V$ can also evaluate the quality of measurement axes chosen with existing methods for sparse signal recovery on the sphere \cite{rauhut2011sparse, alem2012sparse}.
In practice, we find that uniform random sampling outperforms these methods, in the sense that it generates low-cost sets of measurement axes with higher probability.
Two observations may explain the poor performance of these sparse signal recovery methods.
First, spin qudit tomography essentially boils down to recovering sparse signals with a particular harmonic structure, namely with a fixed degree $\ell$.
The existing sparse signal recovery algorithms do not assume or account for any harmonic structure, and may therefore perform worse than expected on this particular subset of sparse signals.
Second, spin qudit tomography requires recovering signals that are band-limited at degree $d$, but have $O(d)$ non-zero harmonic coefficients (out of $d^2$ coefficients in total).
The existing sparse signal recovery methods, on the other hand, may only perform well for signals with sparsity much smaller than $d$.
In any case, the fact that uniform sampling outperforms these methods only simplifies our randomized tomography protocol.

@@

As functions of Euler angle triplets $\v\omega=\p{\alpha_{\v\omega},\beta_{\v\omega},\gamma_{\v\omega}}$, the fundamental (Wigner) rotation matrix elements $D_{mn}^\ell$ and satisfy the orthogonality relations \cite{brown2003rotational}
\begin{align}
  \int_{\Omega_3} \d\Omega_3\p{\v\omega}
  D^{\ell}_{mn}\p{\v\omega}^* D^{\ell'}_{m'n'}\p{\v\omega}
  = \f{8\pi^2}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'} \delta_{nn'},
  \label{eq:orthogonal}
\end{align}
where integration is performed over all $\v\omega\in\Omega_3\simeq[0,2\pi]\times[0,\pi]\times[0,2\pi]$ with integration measure $\d\Omega_3\p{\v\omega}\equiv\d\alpha_{\v\omega} \d\beta_{\v\omega} \d\gamma_{\v\omega} \sin\beta_{\v\omega}$.
This orthogonality relation is a consequence of the fact that, for a fixed degree $\ell$, the matrix
\begin{align}
  D^\ell \equiv \sum_{m,n} D^\ell_{mn} \op{m}{n}
\end{align}
is an irreducible unitary matrix representation of SO(3), where by unitary we mean that
\begin{align}
  \sum_k D^\ell_{km}\p{\v\omega}^* D^\ell_{kn}\p{\v\omega} = \delta_{mn}
  ~\t{for all}~\v\omega\in\Omega_3\simeq\SO(3).
\end{align}
Unitarity of $D^\ell$ follows from the fact that the matrix $D^\ell\p{\v\omega}$ is essentially the degree-$\ell$ block of the block-diagonal fundamental rotation matrix $R\p{\v\omega}$, which is itself unitary.
The Schur orthogonality relations for irreducible unitary representations of compact Lie groups then imply that \cite{cornwell1997group}
\begin{align}
  \int_{\Omega_3} \d\Omega_3\p{\v\omega}
  D^{\ell}_{mn}\p{\v\omega}^* D^{\ell'}_{m'n'}\p{\v\omega}
  = \f{\abs{\Omega_3}}{\dim\p{D^\ell}}
  \delta_{\ell\ell'} \delta_{mm'} \delta_{nn'},
  \label{eq:ortho_full_apndx}
\end{align}
where $\dim\p{D^\ell}=2\ell+1$ is the dimension of $D^\ell$, and analogously
\begin{align}
  \abs{\Omega_3} \equiv \int_{\Omega_3} \d\Omega_3\p{\v\omega} = 8\pi^2.
\end{align}
Orthogonality relations for adjoint matrix elements $\D^\ell_{mn}$ are derived identically from the fact that the matrix
\begin{align}
  \D^\ell \equiv \sum_{m,n} \D^\ell_{mn} \op{m}{n}
\end{align}
is an irreducible unitary representation of SO(3).
This property of $\D^\ell$ follows from the fact that it is an adjoint representation of SO(3), which is isomorphic to the standard representation $D^\ell$ \cite{hall2015lie}.

The decomposition $D^\ell_{mn}\p{\v\omega} = e^{-\i\alpha_{\v\omega}m} d^\ell_{mn}\p{\beta_{\v\omega}} e^{-\i\gamma_{\v\omega}n}$, where $d^\ell_{mn}\p{\beta}\equiv D^\ell_{mn}\p{0,\beta,0}$ is
a matrix element of a reduced fundamental (Wigner) rotation matrix \cite{rose1957elementary}, implies that setting $n'=n$ in in Eq.~\eqref{eq:ortho_full_apndx} makes the integrand independent of $\gamma_{\v\omega}$.
Evaluating the integral over $\gamma_{\v\omega}$ and dividing through by $2\pi$ then recovers the orthogonality relations
\begin{align}
  \int_{\Omega_2} \d\Omega_2\p{\v v}
  D^{\ell}_{mn}\p{\v v}^* D^{\ell'}_{m'n}\p{\v v}
  = \f{4\pi}{2\ell+1} \delta_{\ell\ell'} \delta_{mm'},
  \label{eq:ortho_sphere_apndx}
\end{align}
where integration is performed over all $\v v\in\Omega_2\simeq[0,2\pi]\times[0,\pi]$ with integration measure $\d\Omega_2\p{\v v} \equiv \d\alpha_{\v v} \d\beta_{\v v} \sin\beta_{\v v}$.
A similar decomposition of the adjoint rotation matrix elements $\D^\ell_{mn}\p{\v\omega}$ (see Appendix \ref{sec:rotations}) implies the identical orthogonality relations.

@@

Here we show how one solution to the problem of signal recovery on the sphere with a particular basis of functions (such as the spherical harmonics $Y_{\ell m}$) can be translated into a solution with a different basis (such as the adjoint rotation matrix elements $\D^\ell_{m,0}$).
Consider a square-integrable function on the sphere, $f\in L^2\p{\Omega_2}$, with the harmonic expansion
\begin{align}
  f\p{\v v} = \sum_{\ell,m} f_{\ell m} B_{\ell m},
\end{align}
where $f_{\ell m}\in\CC$ is a coefficient and $\set{B_{\ell m}}$ is a complete basis for $L^2\p{\Omega_2}$.
Without loss of generality, we will assume that for a fixed degree $\ell$ the functions $\set{B_{\ell m}}$ span the degree-$\ell$ subspace of $L^2\p{\Omega_2}$, and that these functions satisfy orthogonality relations analogous to those in Eqs.~\eqref{eq:ortho_sphere} and \eqref{eq:ortho_sphere_apndx}.
A solution to the signal recovery problem amounts to a set of points $V=\set{\v v}\subset\Omega_2$ and coefficients $C_{\ell m}\p{\v v}$ for which
\begin{align}
  f_{\ell m} = \sum_{\v v} C_{\ell m}\p{\v v} f\p{\v v}.
  \label{eq:recovery_solution}
\end{align}
Consider now a different basis $\set{\tilde B_{\ell m}}$ for $L^2\p{\Omega_2}$, satisfying the same general assumptions that we made for $\set{B_{\ell m}}$, with which
\begin{align}
  f\p{\v v} = \sum_{\ell, m} \tilde f_{\ell m} \tilde B_{\ell m}.
\end{align}
Substituting the harmonic expansion of $f\p{\v v}$ in the original basis $\set{B_{\ell m}}$, we find that
\begin{align}
  \tilde f_{\ell m} = \int_{\Omega_2} \d\Omega_2\p{\v v}
  \tilde B_{\ell m}\p{\v v}^* f\p{\v v}
  = \sum_n G^\ell_{mn} f_{\ell n},
  &&
  G^\ell_{mn} \equiv \int_{\Omega_2} \d\Omega_2\p{\v v}
  \tilde B_{\ell m}\p{\v v}^* B_{\ell n}\p{\v v},
  \label{eq:basis_change}
\end{align}
where integration is performed over all $\v v\in\Omega_2\simeq[0,2\pi]\times[0,\pi]$ with integration measure $\d\Omega_2\p{\v v} \equiv \d\alpha_{\v v} \d\beta_{\v v} \sin\beta_{\v v}$.
Note that the inner product between functions $\tilde B_{\ell m}\p{\v v}$ and $B_{kn}\p{\v v}$ for $k\ne\ell$ vanishes because these functions live in orthogonal subspaces of $L^2\p{\Omega_2}$ \cite{freeden2008spherical}.
Substituting Eq.~\eqref{eq:recovery_solution} into Eq.~\eqref{eq:basis_change}, we find that
\begin{align}
  \tilde f_{\ell m}
  = \sum_n G^\ell_{mn} \sum_{\v v} C_{\ell n}\p{\v v} f\p{\v v}
  = \sum_{\v v} \tilde C_{\ell m}\p{\v v} f\p{\v v},
  &&
  \tilde C_{\ell m}\p{\v v} \equiv \sum_n G^\ell_{mn} C_{\ell n}\p{\v v}.
\end{align}
The coefficients $\tilde C_{\ell m}\p{\v v}$ thus solve the signal recovery problem with the new basis $\set{\tilde B_{\ell m}}$ for $L^2\p{\Omega_2}$.
